Re-Examine Hybrid State-Space and Self-Attention for In-Context Learning

### 2025-01-30 paper.tex
re compose type

% Use the following line for the initial blind version submitted for review:
% \usepackage{icml2025}

% If accepted, instead use the following line for the camera-ready submission:
\usepackage[accepted]{icml2025}

@@

\section{Background}
\subsection{State-Space}

@@

\begin{equation}
\begin{aligned}
   h_t &= Ah_{t-1} + Bx_t \\
   y_t &= Ch_t
\end{aligned}
\label{eq:state_space}
\end{equation}
% 从连续时间系统的状态空间模型出发，可以通过离散化方法得到离散时间的状态更新公式,方程 (2a) 和 (2b) 代表状态空间模型的离散化形式。它们描述了在离散时间间隔内，状态向量和输出向量如何基于前一个状态和当前输入进行更新。
% \end{alignat}
Starting from the state-space model of continuous-time systems, the discrete-time state update formula can be obtained through discretization methods. Equations (2a) and (2b) represent the discrete form of the state-space model. They describe how the state vector and output vector are updated based on the previous state and current input at discrete time intervals.

@@

% 通过自注意力机制，Transformer模型能够直接捕捉序列中任意两个元素之间的关系，从而有效处理长距离依赖问题。然而，自注意力机制的二次复杂度限制了Transformer模型的计算效率，使得其在处理长序列数据时表现不佳。为了克服这一局限性，研究者们提出了许多改进方法，例如~\cite{kitaev2020reformer}提出了一种基于局部敏感哈希的自注意力机制，通过减少查询和键之间的点积计算量，从而提高了Transformer模型的计算效率。同样, 这种改进方法仍然无法完全解决Transformer模型的计算效率问题, 因此揭示自注意力机制和状态空间之间的更深层次关系, 并提出一种新的混合建模架构, 以克服这些局限性是非常重要的.

Through the self-attention mechanism, the Transformer model can directly capture the relationship between any two elements in the sequence, effectively handling long-distance dependency problems. However, the quadratic complexity of the self-attention mechanism limits the computational efficiency of the Transformer model, making it perform poorly when handling long sequence data. To overcome this limitation, researchers have proposed many improvement methods. For example, Lin et al.~\cite{kitaev2020reformer} proposed a self-attention mechanism based on local sensitive hashing, which reduces the computational cost of dot products between queries and keys, thereby improving the computational efficiency of the Transformer model. Similarly, this improvement method still cannot completely solve the computational efficiency problem of the Transformer model. Therefore, it is very important to reveal the deeper relationship between the self-attention mechanism and the state space and propose a new hybrid modeling architecture to overcome these limitations.

\subsection{Hybrid Modeling}
% 混合建模架构模型是一种将状态空间和自注意力进行混合建模的模型，其目的是使模型具备与Mamba类似的效率和与Transformer类似的效果。例如，jamba~\cite{lieber2024jamba}模型是一种基于混合建模架构的通用语言模型，其基本结构为8层混合循环，每个混合循环包括7个state-space模块、1个self-attention模块和8个MLP模块。通过将状态空间和自注意力进行混合建模，jamba模型能够在大多数语言建模任务上取得更好的性能。基于混合建模架构的模型在大多数语言建模任务上的表现超越了单独使用Mamba或Transformer，但是在情景学习任务上的表现并不理想。

Hybrid modeling architecture models are models that combine state-space and self-attention for hybrid modeling, with the goal of making the model have efficiency similar to Mamba and effect similar to Transformer. For example, the Jamba~\cite{lieber2024jamba} model is a universal language model based on a hybrid modeling architecture, with a basic structure of 8 layers of hybrid cycles, each hybrid cycle consisting of 7 state-space modules, 1 self-attention module, and 8 MLP modules. By combining state-space and self-attention for hybrid modeling, the Jamba model can achieve better performance on most language modeling tasks. Models based on hybrid modeling architecture outperform using Mamba or Transformer alone in most language modeling tasks, but their performance in in-context learning tasks is not ideal.

@@

@@

\begin{table*}[!htbp]
   \footnotesize
   \centering
   \caption{
      \textbf{Parameters of Jamba and Jamba-SABLH for Associative Recall}.
      % Jamba的混合循环相当于循环表~\ref{tab:Jamba_vs_Jamba-SABLH_Params} 中的混合配置三次, Jamba-SABLH 除了最后一层的 self-attention 模块外, 其余参数设置与 Jamba 相同. 除此之外, 我们还是用混合专家作为前馈网络来减少训练成本. 我们通过仔细地调整 $d_ffn$ 来使参数量保持在 1B.
      The hybrid cycle of Jamba is equivalent to the hybrid configuration in Table~\ref{tab:Jamba_vs_Jamba-SABLH_Params} three times, and Jamba-SABLH is the same as Jamba except for the last layer of the self-attention module. In addition, we use mixture of experts as the feed-forward network to reduce training costs. We carefully adjusted $d_{ffn}$ to keep the parameter volume at 1B.
   }
   \label{tab:associative_recall_params}
   \vskip 0.15in
   \begin{tabular}{@{}llllllllllllllllll@{}}
   \toprule
   \sc{Model} & \sc{Params} & $n_{layers}$ & $d_{model}$ & $n_{heads}$ & $d_{state}$ & $n_{experts}$ & $top_{experts}$ \\
   \midrule
   Jamba & 1B & 24 & 1024 & 16 & 64 & 4 & 1 \\
   Jamba-SABLH & 1B & 24 & 1024 & 16 & 64 & 4 & 1 \\
   \bottomrule
   \end{tabular}
\end{table*}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\appendix
\onecolumn
\section{You \emph{can} have an appendix here.}

You can have as much text here as you want. The main body must be at most $8$ pages long.
For the final version, one more page can be added.
If you want, you can use an appendix like this one.

The $\mathtt{\backslash onecolumn}$ command above can be kept in place if you prefer a one-column appendix, or can be removed if you prefer a two-column appendix.  Apart from this possible change, the style (font size, spacing, margins, page numbering, etc.) should be kept the same as the main body.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

% Use the following line for the initial blind version submitted for review:
\usepackage{icml2025}

% If accepted, instead use the following line for the camera-ready submission:
% \usepackage[accepted]{icml2025}

@@

\section{Background}

\subsection{State-Space}

@@

\begin{equation}
\begin{aligned}
   h_t &= Ah_{t-1} + Bx_t \\
   y_t &= Ch_t
\end{aligned}
\label{eq:state_space}
\end{equation}

% 从连续时间系统的状态空间模型出发，可以通过离散化方法得到离散时间的状态更新公式,方程 (2a) 和 (2b) 代表状态空间模型的离散化形式。它们描述了在离散时间间隔内，状态向量和输出向量如何基于前一个状态和当前输入进行更新。
% \end{alignat}
Starting from the state-space model of continuous-time systems, the discrete-time state update formula can be obtained through discretization methods. Equations (2a) and (2b) represent the discrete form of the state-space model. They describe how the state vector and output vector are updated based on the previous state and current input at discrete time intervals.

@@

% 通过自注意力机制，Transformer模型能够直接捕捉序列中任意两个元素之间的关系，从而有效处理长距离依赖问题。然而，自注意力机制的二次复杂度限制了Transformer模型的计算效率，使得其在处理长序列数据时表现不佳。为了克服这一局限性，研究者们提出了许多改进方法，例如~\cite{kitaev2020reformer}提出了一种基于局部敏感哈希的自注意力机制，通过减少查询和键之间的点积计算量，从而提高了Transformer模型的计算效率。同样, 这种改进方法仍然无法完全解决Transformer模型的计算效率问题, 因此揭示自注意力机制和状态空间之间的更深层次关系, 并提出一种新的混合建模架构, 以克服这些局限性是非常重要的.
Through the self-attention mechanism, the Transformer model can directly capture the relationship between any two elements in the sequence, effectively handling long-distance dependency problems. However, the quadratic complexity of the self-attention mechanism limits the computational efficiency of the Transformer model, making it perform poorly when handling long sequence data. To overcome this limitation, researchers have proposed many improvement methods. For example, Lin et al.~\cite{kitaev2020reformer} proposed a self-attention mechanism based on local sensitive hashing, which reduces the computational cost of dot products between queries and keys, thereby improving the computational efficiency of the Transformer model. Similarly, this improvement method still cannot completely solve the computational efficiency problem of the Transformer model. Therefore, it is very important to reveal the deeper relationship between the self-attention mechanism and the state space and propose a new hybrid modeling architecture to overcome these limitations.

\subsection{Hybrid Modeling}
% 混合建模架构模型是一种将状态空间和自注意力进行混合建模的模型，其目的是使模型具备与Mamba类似的效率和与Transformer类似的效果。例如，jamba~\cite{lieber2024jamba}模型是一种基于混合建模架构的通用语言模型，其基本结构为8层混合循环，每个混合循环包括7个state-space模块、1个self-attention模块和8个MLP模块。通过将状态空间和自注意力进行混合建模，jamba模型能够在大多数语言建模任务上取得更好的性能。基于混合建模架构的模型在大多数语言建模任务上的表现超越了单独使用Mamba或Transformer，但是在情景学习任务上的表现并不理想。
Hybrid modeling architecture models are models that combine state-space and self-attention for hybrid modeling, with the goal of making the model have efficiency similar to Mamba and effect similar to Transformer. For example, the Jamba~\cite{lieber2024jamba} model is a universal language model based on a hybrid modeling architecture, with a basic structure of 8 layers of hybrid cycles, each hybrid cycle consisting of 7 state-space modules, 1 self-attention module, and 8 MLP modules. By combining state-space and self-attention for hybrid modeling, the Jamba model can achieve better performance on most language modeling tasks. Models based on hybrid modeling architecture outperform using Mamba or Transformer alone in most language modeling tasks, but their performance in in-context learning tasks is not ideal.

@@

\begin{table*}[!htbp]
   \centering
   \caption{
      \textbf{Parameters of Jamba and Jamba-SABLH for Associative Recall}.
      % Jamba的混合循环相当于循环表~\ref{tab:Jamba_vs_Jamba-SABLH_Params} 中的混合配置三次, Jamba-SABLH 除了最后一层的 self-attention 模块外, 其余参数设置与 Jamba 相同. 除此之外, 我们还是用混合专家作为前馈网络来减少训练成本. 我们通过仔细地调整 $d_ffn$ 来使参数量保持在 1B.
      The hybrid cycle of Jamba is equivalent to the hybrid configuration in Table~\ref{tab:Jamba_vs_Jamba-SABLH_Params} three times, and Jamba-SABLH is the same as Jamba except for the last layer of the self-attention module. In addition, we use mixture of experts as the feed-forward network to reduce training costs. We carefully adjusted $d_{ffn}$ to keep the parameter volume at 1B.
   }
   \label{tab:associative_recall_params}
   \vskip 0.15in
   \begin{tabular}{@{}llllllllllllllllll@{}}
   \toprule
   \sc{Model} & \sc{Params} & $n_{layers}$ & $d_{model}$ & $n_{heads}$ & $d_{state}$ & $n_{experts}$ & $top_{experts}$ \\
   \midrule
   Jamba & 1B & 24 & 1024 & 16 & 64 & 4 & 1 \\
   Jamba-SABLH & 1B & 24 & 1024 & 16 & 64 & 4 & 1 \\
   \bottomrule
   \end{tabular}
\end{table*}

@@

@@

\section*{Impact Statement}

This paper presents work whose goal is to advance the field of 
Machine Learning. There are many potential societal consequences 
of our work, none which we feel must be specifically highlighted here.

@@
